---
title: "CB_Seminar_Beyer_Homework_02_Elias_Schwall"
author: "Elias Schwall"
date: "`r Sys.Date()`"
output: 
    html_document:
        toc: true
        theme: united
---

```{r echo=TRUE}
# Setting up the linear models 
sepal_length_model_with_predictor_sepal_width <- lm(Sepal.Length ~ Sepal.Width, data = iris)
sepal_length_model_with_predictor_petal_length <- lm(Sepal.Length ~ Petal.Length, data = iris)
sepal_length_model_with_predictor_petal_width <- lm(Sepal.Length ~ Petal.Width, data = iris)

# Looking at the r squared values
summary(sepal_length_model_with_predictor_sepal_width)$r.squared
summary(sepal_length_model_with_predictor_petal_length)$r.squared
summary(sepal_length_model_with_predictor_petal_width)$r.squared

# Looking at the RSS of each model 
sum(resid(sepal_length_model_with_predictor_sepal_width)^2)
sum(resid(sepal_length_model_with_predictor_petal_length)^2)
sum(resid(sepal_length_model_with_predictor_petal_width)^2)
```
When looking at the r squared values, we see that the r squared value of the model with the petal length predictor has the highest value of `r summary(sepal_length_model_with_predictor_petal_length)$r.squared`, while the model with the sepal width as a predictors has the lowest value of `r summary(sepal_length_model_with_predictor_sepal_width)$r.squared`. \newline
Subsequently, the model with the petal length as a predictor has the lowest RSS: `r sum(resid(sepal_length_model_with_predictor_petal_length)^2)` and it therefore the best predictor of sepal length of the 3 variables, while the model with sepal width as a predictor performs the worst in predicting sepal length (`r sum(resid(sepal_length_model_with_predictor_sepal_width)^2)`). 

```{r Plotting petal length predictor model, echo=TRUE}
library(ggplot2)

ggplot(iris, aes(x = Petal.Length, y = Sepal.Length)) +
  geom_smooth() +
  geom_point() +
  ggtitle("Regression line for the petal length predictor model")

```

