---
title: "CompBio Homework SCHWALL ELIAS"
author: "Elias Schwall"
date: "`r Sys.Date()`"
output: pdf_document
---

# Task 1 Pizza and Markov chains

## a) Give the transition matrix to this process
```{r}
transition <- matrix(c(0.4,0.3,0.3,
                       0.0,0.1,0.9,
                       0.1,0.4,0.5), nrow = 3, ncol = 3, byrow = TRUE)

names <- c("Aurora Pizza", "Pizza Boy", "Pizza Company")
rownames(transition) <- names
colnames(transition) <- names
print(transition)
```
## b) Calculate the costumer distribution after 1,2,10 and 20 month
```{r}
time_frames <- c(1,2,10,20)

for (i in time_frames){
    customer_distribution <- c(200,300,100)
    for (j in 1:i) {
        customer_distribution <- customer_distribution %*% transition
    }
    cat("Customer distrinbution after", i, "months",customer_distribution,"\n", sep = " ")
}
```
## c) Interpret the result, using the term ’stationary distribution’
After 10 months we can see that the customer distribution is basically not changing anymore (rounded). This is called the stationary distribution. The stationary distribution is the long term distribution of the customer distribution. This implies that following a specific duration, the customer distribution for each pizza store will reach a stable state, and the market shares will remain relatively consistent over time. In principle the stantionary distribution is achieved earlier than after 10 months, however in the observed time frames we see behavior after 10 months, when comparing it to 20.

# Task 2 Asymmetric random walk - an urn model 
## a) Define the procedure draw and replace in the R-script, which updates the balls in the urn. For weighted sampling, use sample with the option prob= .... The vector can be defined with the weights 1 and s (1 where balls=0 and s where balls=1)
```{r}
#Urn-model#
N <- 100 #number of balls
s <- 2 #weight of white balls

balls <- rep(0,N)
balls[1] <- 1 #start with one white ball

########################################
# Define the function draw and replace #
########################################

draw_and_replace <- function(balls, s=2){
  # Random deletion
  balls <- balls[-sample(length(balls),1)]
  # Weighted duplication
  prob_weights <- balls
  prob_weights[prob_weights == 1] <- s
  prob_weights[prob_weights == 0] <- 1
  selected_ball <- sample(balls, 1, prob = prob_weights)
  balls <- c(balls, selected_ball)
  return(balls)
}

### Try out ###
balls <- draw_and_replace(balls, s=2)  
print(balls)
```
## b) Describe the purpose of the function draw trajectory. Choose at least 3 different parameter combinations (varying s, N, run time etc.) and attach the plots to the pdf. Give a short interpretation of the effects of the parameters.

The draw_trajectory function is used to simulate the urn model and plot the trajectory of the number of balls in the urn over time. The function takes the following parameters: repl is the number of simulations to be run, run_time is the number of time steps to be simulated, s is the weight of the white balls, and N is the number of balls in the urn. The function is used to investigate the effect of the parameters on the urn model. The following plots show the effect of varying s, N, and repl on the urn model.
```{r}
draw_trajectory <- function(repl=20, run_time=1e3, s=2, N=100){
  s <- s
  trajectory <- rep(0,run_time)
  plot(0,0,xlim=c(0,run_time),ylim=c(0,N),xlab="time",ylab="number",main=paste0("s= ",s,", N=",N,", repl= ",repl,", run_time=",run_time,sep=""))
  for (it in 1:repl) {
    balls <- rep(0,N)
    balls[1] <- 1
    
    for (tm in 1:run_time) {
      trajectory[tm] <- sum(balls)
      balls <- draw_and_replace(balls, s)
    }
    
    points(1:run_time,trajectory,type="l")
  }
}
```

### Plots for varying s
```{r}
# Changing s 
draw_trajectory(repl=20, run_time=1e3, s=1, N=100)
draw_trajectory(repl=20, run_time=1e3, s=5, N=100)
draw_trajectory(repl=20, run_time=1e3, s=10, N=100)
```
When changing the weight of the white balls, we can see that the number of white balls in the urn increases faster when the weight is higher. This is because the probability of drawing a white ball is higher when the weight is higher. 
\newpage

### Plots for varying repl
```{r}
# Changing repl
draw_trajectory(repl=20, run_time=1e3, s=2, N=100)
draw_trajectory(repl=30, run_time=1e3, s=2, N=100)
draw_trajectory(repl=50, run_time=1e3, s=2, N=100)
```
When changing the number of simulations, we can see that there are more trajectories. This is because the number of simulations is the number of times the urn model is simulated. Therefore, it more likely that some trajectories will approach the maximum number of white balls in the urn by chance.
\newpage

### Plots for varying N
```{r}
# Changing N
draw_trajectory(repl=20, run_time=1e3, s=2, N=100)
draw_trajectory(repl=20, run_time=1e3, s=2, N=500)
draw_trajectory(repl=20, run_time=1e3, s=2, N=1000)
```
When changing the number of balls in the urn, we can see that it is more likely to pick a black ball, because the ratio of white to black balls changes (we still start only with one white ball). Therefore, it is more likely that the trajectory will approach the minimum number of white balls (0) in the urn by chance. 
\newpage

### Plots for varying run_time
```{r}
# Changing run_time
draw_trajectory(repl=20, run_time=1e2, s=2, N=100)
draw_trajectory(repl=20, run_time=1e3, s=2, N=100)
draw_trajectory(repl=20, run_time=1e4, s=2, N=100)
```
When changing the number of time steps, we can see that reaching a minimum or maximum number of white balls in 


## c) Describe the purpose of the function get trans prob. Choose at least 10 different parameter combinations (varying x0, s, N). Give a short interpretation of your result.

## d) Bonus: How can this process be applied to population genetics and evolution?

# Task 3 CpG islands and HMMs
## a) Compare the entries of C−G− and C+G+ and interprete the result.

## b) How does the emission matrix look like? What are the observables and the emission
probabilities from the hidden states? (1P)

## c) Use the transition matrix to calculate the probability, that the following sequence was sampled from a CpG-island and the probability, that it was sampled from a non-island region. TGACGACGAA. Compare the probabilities.
