---
title: "CB_Seminar_Homework_Beyer_03"
author: "Elias Schwall"
date: "`r Sys.Date()`"
output: html_document
---

### Exercise: Predicting Tissue with gene expression.
load the files “ExpressionPancreasLung1000genes.Rdata” and “Annotation.RData” into your R session. Take time to get familiar with the data. ExpressionPancreasLung1000genes consists of expression levels of 1000 genes (rows) across 673 samples. The annotation of samples to the respective tissue can be found in the “tissue” column of the Annotation object.
```{r}
load("Expression_Pancreas_Lung_1000genes.Rdata")
Expression_Pancreas_Lung_1000genes <- as.data.frame(Expression_Pancreas_Lung_1000genes)
load("Annotation.RData")
Annotation <- as.data.frame(Annotation)
```

##### a) Model tissue with each of the 1000 genes using generalized linear model. For each model retrieve p-value of the coefficient that shows effect of gene on tissue, using coefficients(summary()). How many genes are predictive of tissue, at a significance threshold of 0.05?
```{r warning=FALSE}
# Creating empty vector for p-values
p_values <- c()

# Creating a model for each gene
for (i in 1:1000){
  model <- glm(tissue ~ Expression_Pancreas_Lung_1000genes[,i], 
               data = Annotation, 
               family = "binomial")
  
 p_values[i] <- summary(model)$coefficients[2,4]
}

# How many genes are predictive of tissue, at a significance threshold of 0.05?
p_values_sig <- dplyr::filter(as.data.frame(p_values), p_values < 0.05)
```
There are `r nrow(p_values_sig)` genes predictive of tissue at a significance threshold of 0.05.

<br>
<br>

##### b) Apply multiple hypothesis testing to the p-values you obtained. See ?p.adjust. How many genes are now predictive of tissues at a significance threshold of 0.05?
```{r}
p_values <- unlist(p_values)
p_values <- as.numeric(p_values)
# Applying multiple hypothesis testing to the p-values
p_values_adj <- p.adjust(p_values, method = "BH")
p_values_adj <- dplyr::filter(as.data.frame(p_values_adj), p_values_adj < 0.05)
```

After adjusting the p-values we are left with `r nrow(p_values_adj)` genes predictive of tissue at a significance threshold of 0.05.
